---
title: "Part 1 Analysis Script"
author: "Luisa M Mimmi"
date: "2/20/2018"
output:
  html_document:
    number_sections: yes
    toc: yes
  word_document:
    toc: yes
---

#	Introduction
I will explore the indicator **_Life expectancy at birth_** obtained from *WDI* to understand how it changed over time (1960 - 2015) and across countries. 
I also will do some very quick exploration of other indicators that I expect to be highly correlated with life expectancy: namely **GDP** and the status of **Fragile and conflict affected situations** (FCS). 

+ In order to flag the individual countries that are classified as FCS, I used the World Bank list of economies (as of June 2017) found [here](http://databank.worldbank.org/data/download/site-content/CLASS.xls)


#	Pull data from an API
```{r packages , results='hide',  message = FALSE}
library(WDI)
library(wbstats)
library(ggplot2) 
library(dplyr)
library(stats)
library(tidyr)
library(stringr)
library(gridExtra)
library(readr)  # for read_csv
library(knitr)  # for kable
```

### Find out IDs for interesting indicators using `WDIsearch` and `wbsearch`
```{r search,  results='hide'}
WDIsearch("life expectancy at birth.*total")
WDIsearch("life expectancy at birth.*male")
wbsearch("gdp.*capita.*US\\$") 
```

### Download those data
```{r data,  results='hide', warning=FALSE,  message = FALSE}
# WDI life expectancy indicators 
WDIlifeexp <- WDI(country = "all", indicator = c("SP.DYN.LE00.IN","SP.DYN.LE00.MA.IN","SP.DYN.LE00.FE.IN", "SI.SPR.PCAP"), start=1960, end=2016, extra= TRUE)
names(WDIlifeexp)

# wb GDP  per capita (wb does not have arg extra)
wbGDP <- wb(country = "all", indicator = c("NY.GDP.PCAP.KD","NY.GDP.PCAP.CD"))
names(wbGDP)
```



#	Basic data wrangling tasks
### Check comparability of *life expectancy* and *gdp* data 

From a quick check, I see that the two tables have different number of countries and that changes over different years. I will give priority to the data of life expectancy which is my outcome of interest. 
```{r check,  results='hide', warning=FALSE,  message = FALSE}
n_distinct(wbGDP$iso2c)      # 255 obs = countries + aggreagates
n_distinct(WDIlifeexp$iso2c) # 264 obs = countries + aggreagates

```
### Merge gdp and life exp then remove "aggregates"
The dataframe `wbGDP` has the 2 GDP indicators in the same column and I want them in 2 columns for consistency 
```{r merge GDP,  results='hide', warning=FALSE,  message = FALSE}
# tidy GDP spread 
wbGDP$date = as.numeric(wbGDP$date)
wbGDP_spread <- select (wbGDP, -indicator ) %>% spread(indicatorID, value, fill = NA, convert = TRUE, drop = TRUE,  sep = NULL) 

# rename GDP
names(wbGDP_spread)[5]<-"GdpPerCapCurrent"
names(wbGDP_spread)[6]<-"GdpPerCap2010"

# merge life and GDP (spread)
lifemerged <-  left_join(WDIlifeexp, wbGDP_spread, by=c("year"= "date", "iso2c"="iso2c")) # 15,048
```


### Additional merge with table to flag *FCS* individual countries 
As I am interested in the relationship between a "Fragile and Conflic Status" and *life expectancy*, but I could not find a WDI variable to flag the individual FCS countries, I merge my dataset with the table **_"FCS.csv" _** I found on line.
```{r merge FCS ,  results='hide', warning=FALSE,  message = FALSE}
# add FCS,csv in a way that is reproducible
# FCS <- read.csv("FCS.csv", header = TRUE, sep = ",") 
myfile <- "https://raw.githubusercontent.com/Lulliter/DECdatascience/master/FCS.csv"
FCS <- read_csv(myfile)

# Merge
lifemergedFCS <- left_join(lifemerged, FCS, by =c("iso3c.x" = "CountryCode")) 
head(lifemergedFCS)

# remove region == Aggreagates -> from 15,048 obs to 12,027 obs 
lifeCountry <- filter(lifemergedFCS, region!= "Aggregates") %>% select(year, "iso2c", country="country.x", "region", incomeLev="income", fragile="GroupCode", lifeexpeALL="SP.DYN.LE00.IN", lifeexpeMA="SP.DYN.LE00.MA.IN",   lifeexpeFE="SP.DYN.LE00.FE.IN",  starts_with("GdpP"))
  
summary(lifeCountry)
n_distinct(lifeCountry$iso2c)    # 304 obs = countries (211) + "aggreagates"
n_distinct(lifeCountry$country)  #  countries (211) + "aggreagates"
n_distinct(lifeCountry$region)   #  7 + "aggreagates"
n_distinct(lifeCountry$incomeLev) #  6 + "aggreagates"




```



#	Descriptive Statistics considerations
###  Dealing with missing observations for life expectancy indicators  
After checking across the key dimensions of my dataset, I observe that missing values have a very high incidence among small countries/islands. Since this seems to be a coverage error in the sample that should not affect the outcome of interest, I decide it is OK to exclude them for the purpose of this analysis. I obtain 10,743 instead of 12,027 observations
```{r missing values ,  results='hide', warning=FALSE,  message = FALSE}
# Count NA obs
sum(is.na(lifeCountry$lifeexpeALL)) # 1284 
sum(is.na(lifeCountry$lifeexpeMA)) # 1284 
sum(is.na(lifeCountry$lifeexpeFE)) # 1284 
sum(is.na(lifeCountry$GdpPerCapCurrent)) # 2965
sum(is.na(lifeCountry$GdpPerCap2010)) # 3399

# Explore if NA distribution is sistematically related to any key dimensions
table (lifeCountry$country , is.na(lifeCountry$lifeexpeALL))  # looks like small countries are the ones missing
table (lifeCountry$region , is.na(lifeCountry$lifeexpeALL)) # not too much difference across regions 
table (lifeCountry$fragile , is.na(lifeCountry$lifeexpeALL)) # similar incidence as other groups 
table (lifeCountry$year , is.na(lifeCountry$lifeexpeALL))  # year 2015 is the last available 
table (lifeCountry$incomeLev , is.na(lifeCountry$lifeexpeALL)) # notable high incidenc eof missing in "High income: nonOECD""

# Drop missing "lifeexpALL" 
lifeCountry2 <- lifeCountry[!is.na(lifeCountry$lifeexpeALL),]

# I recode FCS as dummy variable because NA is actually "0" there
lifeCountry2 <- mutate(lifeCountry2, fragile =  ifelse(is.na(lifeCountry2$fragile), 0, 1))
```

###  Checking correlation between life expectancy and GDP and variation across groups of different income levels.

+ Correlation (measured as Pearson's) between life expectancy and GDP is very high as suspected (0.78 for GDP constant 2010)
+ I also use ANOVA (analisys of variance) to validate whether the mean outcomes differ across groups.The highly significance F test statistics suggests extremely high variation across income levels. 

```{r GDP and income,   warning=FALSE,  message = FALSE}
# 1st transform GDP indicators taking the natural-log for easier change interpretation
lifeCountry2$logGDPcurr <- log(lifeCountry2$GdpPerCapCurrent)
lifeCountry2$logGDPconst <- log(lifeCountry2$GdpPerCap2010)

# 2nd I look at Pearson's correlation
cor(lifeCountry2$lifeexpeALL, lifeCountry2$logGDPconst, use="complete.obs") # 0.7815843
cor(lifeCountry2$lifeexpeALL, lifeCountry2$logGDPcurr, use="complete.obs") # 0.836471

# 2-way ANOVA with factorial A (year) and B (income)
fit2 <- aov(lifeexpeALL ~ as.factor(year) + as.factor(incomeLev) , data=lifeCountry2)
summary(fit2) # display Type I sequential SS (effect as the factor were considered one at a time into the model, in the order they are entered in the model )
drop1(fit2,~.,test="F") # type III marginal SS and F Tests (effect of each variable is evaluated after all other factors have been accounted for)
```


#	Further information through data visualization

### Plot of correlation between life expectancy and GDP
The graph corroborates the high correlation found
```{r g1, echo=FALSE}
g1 <- ggplot (lifeCountry2, aes(x=logGDPconst, y = lifeexpeALL, color=lifeexpeALL)) + geom_point() +   scale_color_gradient(low= "red",high = "green") + labs(title="Correlation of Life Expectancy (All) with log of GDP (constant 2010 value)", x="log of GDP", y="years") 

g1
```


### Comparing means of life expectancy across Income levels in key years
Looking only at 1960 and 2015, I also find striking visual evidence that being a FCS country has consistently lowered life expectancy across almost evey group.
```{r , echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=8}
# plotting vs incomeLev & fragile 
lifeCountry2$incomeLev = str_wrap(lifeCountry2$incomeLev, width = 10)

# reorder income lev
lifeCountry2$incomeLev2 <- factor(lifeCountry2$incomeLev, levels = c("Not\nclassified","Low income","Lower\nmiddle\nincome","Upper\nmiddle\nincome","High\nincome:\nnonOECD","High\nincome:\nOECD"))


le1960 <- lifeCountry2 %>% filter(year==1960) 
g2 <-ggplot( le1960, aes(incomeLev2, lifeexpeALL)) + geom_boxplot((aes(fill=factor(fragile)))) + 
   scale_fill_manual(values=c("#00BA38", "#F8766D" )) +
    theme(axis.text.x = element_text(angle=65, vjust=0.7)) + 
 # scale_x_discrete(labels = abbreviate) +
    labs(title="Life Expectancy (All) by incomeLev2", 
         subtitle="1960",
         x="incomeLev2",
         y="years")
          # + coord_flip()

le2015 <- lifeCountry2 %>% filter(year==2015) 
g3 <- ggplot( le2015, aes(incomeLev2, lifeexpeALL)) + geom_boxplot((aes(fill=factor(fragile)))) + 
   scale_fill_manual(values=c("#00BA38", "#F8766D" )) +
    theme(axis.text.x = element_text(angle=65, vjust=0.7)) + 
 # scale_x_discrete(labels = abbreviate) +
    labs(title="Life Expectancy (All) by incomeLev2", 
         subtitle="2015",
        
         x="incomeLev2",
         y="years")  
  # + coord_flip()

grid.arrange(g2, g3, nrow=2)
```


### Comparing means of life expectancy across Regions in key years
I find again striking visual evidence that being a FCS country lower the life expectancy variable. Even more remarkable when looking at the change in MENA between 1960 and 2015... the divergence between FCS and Not-FCS has increased visibly. South Asia shows the most remarkable improvement of average life expectancy over time.

```{r , echo=FALSE, message=FALSE, warning=FALSE,  fig.width=8, fig.height=8}
# plotting vs region & fragile 
lifeCountry2$region = str_wrap(lifeCountry2$region, width = 15)

le1960 <- lifeCountry2 %>% filter(year==1960) 

g3 <-ggplot( le1960, aes(region, lifeexpeALL)) + geom_boxplot((aes(fill=factor(fragile)))) + 
   scale_fill_manual(values=c("#00BA38", "#F8766D" )) +
    theme(axis.text.x = element_text(angle=65, vjust=0.7)) + 
 # scale_x_discrete(labels = abbreviate) +
    labs(title="Life Expectancy (All) by region", 
         subtitle="1960",
        
         x="region",
         y="years")
          # + coord_flip()

le2015 <- lifeCountry2 %>% filter(year==2015) 


g4 <- ggplot( le2015, aes(region, lifeexpeALL)) + geom_boxplot((aes(fill=factor(fragile)))) + 
   scale_fill_manual(values=c("#00BA38", "#F8766D" )) +
    theme(axis.text.x = element_text(angle=65, vjust=0.7)) + 
 # scale_x_discrete(labels = abbreviate) +
    labs(title="Life Expectancy (All) by region", 
         subtitle="2015",
        
         x="region",
         y="years")  
  # + coord_flip()

grid.arrange(g3, g4, nrow=2)


  #tapply(le1960$lifeexpeFE, le1960$incomeLev, mean)
 #tapply(le2015$lifeexpeFE, le2015$incomeLev, mean)
 #tapply(le2015$lifeexpeMA, le2015$incomeLev, mean)
```

### Patterns of change of life expectancy over time - across Income Level
Upon observing a clear difference in terms of life expectancy distribution across income levels observed in 2 extreme years, I will check the change over time.
Interestingly, in high income countries the lif expectancy of women is consistently above the one of men through the decades. 

```{r , echo=FALSE, fig.width=8, fig.height=8}
# group by income lev
lifeCountry3 <- lifeCountry2 %>% group_by(year, incomeLev2) %>% summarise(meanLEall=mean(lifeexpeALL), meanLEmale=mean(lifeexpeMA), meanLEfem=mean(lifeexpeFE)) 


g5 <- ggplot(lifeCountry3, aes(x=year) ) + 
        geom_line(aes(y=meanLEfem,color="Female Life Expectancy")) +
        scale_colour_discrete(name ="Groups")+
        facet_wrap(~ incomeLev2)+
         geom_line(aes(y=meanLEmale,color="Male Life Expectancy"))+
         geom_line(aes(y=meanLEall,color="Total Life Expectancy")) +
        labs(x="Year",y="life expectancy") +
        ggtitle("Life Expectancy changes across income levels between 1960 and 2015")

g5 
```



### Patterns of change of life expectancy over time - across Region
Upon demonstrating the significant difference of life expectancy's means & trend patterns across Income Level, I want to compare growth patterns across regions 
In terms of historical trends and change patterns, it seems like the region "categorical variable" picks up a much more impressive divergence
```{r ,fig.width=8, fig.height=8}
# have to group by region
lifeCountry4 <- lifeCountry2 %>% group_by(year, region) %>% summarise(meanLEall=mean(lifeexpeALL), meanLEmale=mean(lifeexpeMA), meanLEfem=mean(lifeexpeFE)) 

g6 <- ggplot(lifeCountry4, aes(x=year) ) + 
        geom_line(aes(y=meanLEfem,color="Female Life Expectancy")) +
        scale_colour_discrete(name ="Groups")+
        facet_wrap(~ region)+
         geom_line(aes(y=meanLEmale,color="Male Life Expectancy"))+
         geom_line(aes(y=meanLEall,color="Total Life Expectancy")) +
        labs(x="Year",y="life expectancy in years") +
        ggtitle("Life Expectancy changes across regions between 1960 and 2015")

g6 
```